---
title: "DESeq2 Analysis"
author: "fgaudilliere"
date: "16/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# First analysis: all libraries

## Adapting the salmon output to DESeq2

Loading tximport:
```{r}
library("tximport")
```


Creating a vector with the paths to the different salmon ouputs (forward and reverse in my case because I ran salmon in single-end, so I have separated files for the forward and reverse reads of each library) and naming the columns of the vector I created:
```{r}
data = "/ifb/data/mydatalocal/data_tp_ngs"
file_names = c()
for (i in c(1,2,3,4,5,6)) {
  file_names = c(file_names,paste0(data,"/","salmon_alignment_single_end",i,"_forward","/","quant.sf"))
  file_names = c(file_names,paste0(data,"/","salmon_alignment_single_end",i,"_reverse","/","quant.sf"))
}
names(file_names) <- c("Lib1_forward","Lib1_reverse","Lib2_forward","Lib2_reverse","Lib3_forward","Lib3_reverse","Lib4_forward","Lib4_reverse","Lib5_forward","Lib5_reverse","Lib6_forward","Lib6_reverse")
```


Retrieving the trinity file containing the correspondence between trinity gene ID and trinity isoform ID ("trinity"), naming the columns and and inverting the order of the columns ("trinity1"):
```{r}
trinity <- read.table("/ifb/data/mydatalocal/data_tp_ngs/trinity_results/Trinity_RF.fasta.gene_trans_map")
names(trinity) <- c("GENEID","TXNAME")
trinity1 <- trinity[,c(2,1)]
```


Creating a table containing the NumReads value, pooled by gene:
```{r}
tximport_file <- tximport(files = file_names, type = "salmon", tx2gene = trinity1)
```

Loading the DESeq2 library:
```{r}
library("DESeq2")
```

Preparing the import by generating a table indicating, for each library, what the condition is (CTL or INF):
```{r}
samp.name <- names(file_names)
samp.type <- c("CTL","CTL","CTL","CTL","CTL","CTL","INF","INF","INF","INF","INF","INF")
samples <- data.frame(run=samp.name,condition=samp.type)
```


Importing the data from the tximport file into a file adapted to DESeq2:
```{r}
ddsTxi <- DESeqDataSetFromTximport(tximport_file,
                                   colData = samples,
                                   design = ~ condition)
```


Running DESeq2: 
```{r}
dds <- DESeq(ddsTxi)
res <- results(dds)
res
```

Looking at how many genes display a p-value < 0.05 or 0.01: 
```{r}
table(res$padj < 0.05)
table(res$padj < 0.01)
```


Plotting a MA-plot for quality control: 
```{r}
plotMA(res, ylim=c(-2,2))
```

```{r}
vsd <- vst(dds, blind=FALSE)
```

Plotting the PCA for quality control (either by condition or condition + library number):
```{r}
plotPCA(vsd, intgroup=c("condition"))
plotPCA(vsd, intgroup=c("condition", "run"))
```


# Re-analysis without Library 3

Creating a vector with the paths to the different salmon ouputs and naming the columns of the vector I created:
```{r}
data = "/ifb/data/mydatalocal/data_tp_ngs"
file_names2 = c()
for (i in c(1,2,4,5,6)) {
  file_names2 = c(file_names2,paste0(data,"/","salmon_alignment_single_end",i,"_forward","/","quant.sf"))
  file_names2 = c(file_names2,paste0(data,"/","salmon_alignment_single_end",i,"_reverse","/","quant.sf"))
}
names(file_names2) <- c("Lib1_forward","Lib1_reverse","Lib2_forward","Lib2_reverse","Lib4_forward","Lib4_reverse","Lib5_forward","Lib5_reverse","Lib6_forward","Lib6_reverse")
```

Creating a table containing the NumReads value, pooled by gene:
```{r}
tximport_file2 <- tximport(files = file_names2, type = "salmon", tx2gene = trinity1)
```

Preparing the import by generating a table indicating, for each library, what the condition is (CTL or INF):
```{r}
samp.name2 <- names(file_names2)
samp.type2 <- c("CTL","CTL","CTL","CTL","INF","INF","INF","INF","INF","INF")
samples2 <- data.frame(run=samp.name2,condition=samp.type2)
```

Importing the data from the tximport file into a file adapted to DESeq2:
```{r}
ddsTxi2 <- DESeqDataSetFromTximport(tximport_file2,
                                   colData = samples2,
                                   design = ~ condition)
```


Running DESeq2: 
```{r}
dds2 <- DESeq(ddsTxi2)
res2 <- results(dds2)
res2
```

Looking at how many genes display a p-value < 0.05 or 0.01:
```{r}
table(res2$padj < 0.05)
table(res2$padj < 0.01)
```

Plotting a MA-plot for quality control: 
```{r}
plotMA(res2, ylim=c(-2,2))
```


```{r}
vsd2 <- vst(dds2, blind=FALSE)
```

Plotting the PCA for quality control (either by condition or condition + library number):
```{r}
plotPCA(vsd2, intgroup=c("condition"))
plotPCA(vsd2, intgroup=c("condition", "run"))
```





# Merging DESeq2 output with blastn results

Loading blastn output:
```{r}
blast <- read.table(paste0(data,"/blast_alignment/blast_alignment_output"))
```

Loading necessary libraries: 
```{r}
library(dplyr)
library(tidyr)
library(data.table)
library(ggplot2)
library(biomaRt)
```


Creating a DESeq2 results dataframe that includes the Ensembl gene IDs and the gene names instead of only the Trinity gene IDs:
```{r}
# match gene/transcript names
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
tx2geneHomo = biomaRt::getBM(attributes=c('ensembl_transcript_id', 
                                      'ensembl_gene_id', 'external_gene_name'),mart=ensembl)

# add gene name to the blast outputs. we can probably do much better!!
blastHomo<-data.table(blast)
blastHomo<-blastHomo[, c("trid", "version") := tstrsplit(V2, ".", fixed=TRUE)][]
blastHomo<-blastHomo[, c("gene", "alttr") := tstrsplit(V1, "_i", fixed=TRUE)][]
blastHomo1 <- blastHomo%>% group_by(gene)%>% slice_max(order_by = V12, n = 1, with_ties = TRUE)
blastHomoName<-merge(tx2geneHomo,blastHomo1,by.x="ensembl_transcript_id",by.y="trid")
blastHomoNameUniq<-unique(blastHomoName[,c("ensembl_gene_id","external_gene_name","gene")])
# associate our DE results with gene names from human
resUsName=merge(as.data.frame(res2),blastHomoNameUniq,by.x=0,by.y="gene")
```

Creating sub-tables containing the upregulated genes, the downregulated genes and both up- and downregulated genes:
```{r}
UpReg <- filter(resUsName, padj<0.05 & resUsName$log2FoldChange>0)
DownReg <- filter(resUsName, padj<0.05 & resUsName$log2FoldChange<0)
AllReg <- filter(resUsName, padj<0.05)
```


Loading data from Holzer et al. for comparison:
```{r}
resBatHolzer<-read.csv("https://www.rna.uni-jena.de/supplements/mda/data/deseq2/RNA/Mock_6h_RNA_vs_IFN_6h_RNA/deseq2_Mock_6h_RNA_IFN_6h_RNA_full.csv")
```


Merging our results and the Holzer results by gene names:
```{r}
names(resUsName)[9] <- 'gene_name' #renaming a column so that the "gene name" column has the same name in both tables
CommonGenesUsHolze <- merge(resUsName,resBatHolzer,by="gene_name") 
```


Preparing lists of genes in csv format to use in Gorilla:
```{r}
all_genes_list <- resUsName$gene_name
up_reg_list <- UpReg$gene_name
down_reg_list <- DownReg$gene_name
all_reg <- AllReg$gene

write.csv(as.data.frame(unique(all_genes_list)), 
          file="/ifb/data/mydatalocal/data_tp_ngs/gorilla_lists/names_all.csv",quote=FALSE,row.names = FALSE)

write.csv(as.data.frame(unique(up_reg_list)), 
          file="/ifb/data/mydatalocal/data_tp_ngs/gorilla_lists/names_up.csv",quote=FALSE,row.names = FALSE)

write.csv(as.data.frame(unique(down_reg_list)), 
          file="/ifb/data/mydatalocal/data_tp_ngs/gorilla_lists/names_down.csv",quote=FALSE,row.names = FALSE)

write.csv(as.data.frame(unique(all_reg)), 
          file="/ifb/data/mydatalocal/data_tp_ngs/gorilla_lists/names_allreg.csv",quote=FALSE,row.names = FALSE)
```




Testing a heatmap (10 first genes):
```{r}
select <- na.omit(resUsName$Row.names[resUsName$padj<0.05 & resUsName$log2FoldChange>0])
select <-select[1:10]
row.names(samples2)=samples2$run
annot_corresp <- data.frame(resUsName$Row.names,resUsName$gene_name)
colnames(annot_corresp) <- c("nameTrinity","genename")
annot_corresp1 <- filter(annot_corresp, annot_corresp$nameTrinity %in% select)

pheatmap(assay(vsd2)[select,], cluster_rows=TRUE, show_rownames=TRUE,show_colnames=FALSE,
         cluster_cols=FALSE, annotation_col=samples2, annotation_row = annot_corresp1)
```


Creating a column in the merged table to hold the result of the DESeq2 analysis in our experiment (either non-regulated, upregulated or downregulated):
```{r}
CommonGenesUsHolze$ourTest="nonreg"
CommonGenesUsHolze$ourTest[CommonGenesUsHolze$padj.x<0.05&CommonGenesUsHolze$log2FoldChange.x>0]="Up"
CommonGenesUsHolze$ourTest[CommonGenesUsHolze$padj.x<0.05&CommonGenesUsHolze$log2FoldChange.x<0]="Down"
```

Creating a column in the merged table to hold the result of the DESeq2 analysis in the Holzer experiment:
```{r}
CommonGenesUsHolze$theirTest="nonreg"
CommonGenesUsHolze$theirTest[CommonGenesUsHolze$padj.y<0.05&CommonGenesUsHolze$log2FoldChange.y>0]="Up"
CommonGenesUsHolze$theirTest[CommonGenesUsHolze$padj.y<0.05&CommonGenesUsHolze$log2FoldChange.y<0]="Down"
```

Testing whether the overlap between the two experiments in each category is higher than expected by chance:
```{r}
chisq.test(table(us=CommonGenesUsHolze$ourTest,them=CommonGenesUsHolze$theirTest))
```


